Step-by-Step: Training IMU-based Gestures with Live Feedback

### 2024-02-21 docs/latex-nime/nime-template.tex
edited paper for nime

\documentclass{article}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables
\usepackage{hyperref}
\newcommand{\theHalgorithm}{\arabic{algorithm}}
\usepackage[accepted]{mlsys2024}

\mlsystitlerunning{Step-by-Step: Training IMU-based Gestures with Live Feedback}

\begin{document}
\twocolumn[
\mlsystitle{Step-by-Step: Training IMU-based Gestures with Live Feedback}

\begin{mlsysauthorlist}
\mlsysauthor{Michael Schnebly}{h}
\end{mlsysauthorlist}

\mlsysaffiliation{h}{School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, USA}
\mlsyscorrespondingauthor{Michael Schnebly}{michael\_schnebly@g.harvard.edu}

\mlsyskeywords{Machine Learning, MLSys}

\vskip 0.3in

\begin{abstract}
Recognizing user-defined gestures in inertial measurement unit (IMU) data unlocks new forms of creativity and accessibility in human-computer interaction. However, training gesture recognition models is a difficult task that requires a deep understanding of machine learning. We present Step-by-Step, a software tool that allows users to train gesture recognition models with live audiovisual feedback. Step-by-Step uses a simple neural network to learn to recognize and distinguish multiple gestures in IMU timeseries data. Users can train the model by performing gestures and receiving live feedback on the model's performance. Step-by-Step is designed to be accessible to users with no machine learning experience, while providing a powerful codebase for advanced users.
\end{abstract}
]

\printAffiliationsAndNotice{}

\section{Introduction}
For the past 40 years, the keyboard and mouse have been the primary means of human-computer interaction. The reliability and precision of these devices is undisputed, but it comes at a cost. The limited degrees of freedom available in keyboard and mouse inputs dictate which human body movements can be used to control computers, limiting both accessibility and creativity.

The ideal human-computer interface can adapt to a user's needs or preferences and can be used by anyone, regardless of their physical abilities. The challenge in developing such a device is that it must be able to learn, recognize, and distinguish the user's movements. This is a difficult task that requires a deep understanding of machine learning.

In this paper, we present Step-by-Step, a software tool that allows users to train gesture recognition models with live audiovisual feedback. Step-by-Step uses a simple neural network to learn to recognize and distinguish multiple gestures in IMU timeseries data. Users can train the model by performing gestures and receiving live feedback on the model's performance. Step-by-Step is designed to be accessible to users with no machine learning experience, while providing a powerful codebase for advanced users.

\section{Background}
\subsection{Gesture Recognition}
Gesture recognition is the process of recognizing and distinguishing gestures in data. Gestures are body movement patterns that convey information. In a way, keyboards and mice perform gesture recognition by translating body movements into computer inputs. However, these devices are limited to a small set of predefined gestures. The goal of this work is to allow users to define their own gestures and use them to control computers.

@@

\subsection{Neural Networks}
Neural networks are a class of machine learning models that can be used to analyze IMU timeseries and identify patterns such as gestures. Neural networks are composed of layers of "neurons", each of which takes in a set of inputs, performs a simple function, and produces an output. The outputs of one layer are used as inputs to the next. The final layer represents the model's prediction.

Neural networks are trained by providing them with examples of inputs and corresponding correct outputs. The model adjusts its parameters to minimize the difference between its predictions and the groundtruth outputs. This process is called gradient descent and is performed iteratively until the model converges on a solution. Once trained, the model can be used to make predictions on new inputs.

IMUs and neural networks are well-established and have been used together in a wide range of applications for characterizing body movements. However, a machine learning system that can be easily trained to recognize user-defined gestures with performance rivaling that of the mouse and keyboard remains elusive. The goal of this work is to lay the foundation for such a system.

@@

\begin{figure}[h]
    \includegraphics[width=0.5\textwidth]{../figures/model_diagram.jpg}
    \caption{Neural Network Architecture. The sensor branch (left) receives 20 frames (200 ms at 100 Hz) of 3-channel (x, y, z) IMU data. The sensor branch feeds into a 1D-convolutional layer with 8 filters, each also 20 frames in width. The memory branch (right) receives 15 frames of 1-hot encoded prediction data. The two branches are concatenated and fed into a fully connected output layer to predict the currently active gesture.}
    \label{fig:your_label}
\end{figure}

@@

\begin{figure*}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{../figures/realtime_training.jpg}
    \caption{Live Learning Feedback: The user performs a gesture and receives live feedback on the model's performance. Bottom Panel: 10 seconds of training data. Acceleration components (x, y, z) are shown in color; magnitude is shown in grey. Each magnitude peak is marked by a thin vertical line, indicating it has been labelled as a gesture in the training data. Top Panel: 10 seconds of inference data. At t = 0 seconds the neural network is completely untrained. Naive to what is and is not a gesture, it predicts gesture and non-gesture with equal probability (0.5). As the model trains on non-gesture date, it begins to predict non-gesture with higher probability. At t = 5 seconds, the model has converged on a solution that reliably distinguishes non-gesture. At t = 10 seconds, the model has converged on a solution that reliably distinguishes gestures. Given more training data and time to learn, the model's performance would continue to improve with gessture and non-gesture probabilities approaching 1 and 0, respectively. See supplementary video for live demo in the github repo.}
    \label{fig:realtime_training}
\end{figure*}

@@

The current hardware system uses a microcontroller to stream IMU data to a computer for processing. However, given the small size of the model (default settings: ~2kb with ~500 params) and recent advances in embedded machine learning, it should be possible to run the entire system in realtime on a wireless, wearable microcontroller, including labelling, training, and inference. This would allow the system to be used in a wider range of applications, including those where a computer is not available, desirable, or practical. To this end, the authors are currently exploring the possiblity of using Espressifs ESP32-S3 chip, which includes a neural network accelerator and is capable of running Tensorflow Lite Micro.

@@

\section{Closing Remarks}
Step-by-Step serves as a proof-of-concept for realtime learning of user-defined gestures in IMU data. It is worth noting that the strategies employed here are not limited to IMU data and could be applied to data from other sensors. The authors hope that this work will inspire others to explore the possibilities of realtime machine learning in human-computer interaction.

\section{Acknowledgements}
The authors would like the thank Harvard's TinyML course staff for their guidance and support throughout the semester, including: Jason Jabbour, Ike Uchendu, Jessica Quaye, Matthew Stewart, and Vijay Janapa Reddi.

\section{Permissions}
This is unpublished work. All group members collectively do not agree to allow the publication of our final report video on YouTube by the course staff.

% \bibliography{example_paper}
% \bibliographystyle{mlsys2024}

\end{document}

>>>>

\documentclass{nime-alternate} % Uncomment when publishing final version

% Uncomment only one of the ones below
% \usepackage{anonymize} 		   %Uncomment this line to publish
\usepackage[blind]{anonymize}%Uncomment this line for blind review

% Package that enables the use of accents and non 
% standard characters
\usepackage[utf8]{inputenc}

\begin{document}

\conferenceinfo{NIME'24,}{4--6 September, Utrecht, The Netherlands.}
\title{Step-by-Step: Training IMU-based Gestures with Live Feedback}

\label{key}

\numberofauthors{1}

\author{
% 1st. author
\alignauthor
\anonymize{Michael Schnebly}\\
       \affaddr{\anonymize{School of Engineering and Applied Sciences}}\\
       \affaddr{\anonymize{Harvard University}}\\
       \affaddr{\anonymize{Cambridge, MA, USA}}\\
       \email{\anonymize{michael\_schnebly@g.harvard.edu}}
}
% For your initial submission you MUST ANONYMIZE the authors.

\maketitle
\begin{abstract}
This document serves as an application to NIME's 2024 Student Consortium. Within it, I present my interest in joining the NIME research community, as well as my preliminary results for \textit{Step-by-Step}, a software tool that allows users to train gesture recognition models with live audiovisual feedback. \textit{Step-by-Step} uses a simple neural network to learn to recognize and distinguish multiple gestures in inertial measurement unit (IMU) timeseries data. Users train the model by repeatedly performing a target gesture while receiving live feedback on the model’s ability to recognize that gesture. Once satisfactory results are achieved, the user can save the model and deploy it for realtime musical applications, as demonstrated in supplementary live performance videos. An anonymized code repository is available at \url{https://anonymous.4open.science/r/step-by-step-7506/}
\end{abstract} 
\keywords{Gesture Recognition, IMU, Real-time, Neural Network, Training, Inference, Live Feedback}
\ccsdesc[300]{Applied computing~Sound and music computing}
\ccsdesc[300]{Human-centered computing~Gestural input}
\ccsdesc[300]{Machine learning~Neural networks}

% this line creates the CCS Concepts section.
\printccsdesc

\section{Expected Outcomes}

\section{Introduction}
The broad range of possible mappings between body movement and sound afforded by gesture-based digital musical instruments presents the musician with a creative challenge: Which body movements? Which sounds? And, what is the mapping between them \cite{mitchell:mapping}?

A common approach to this challenge are \textit{learned} mappings, for which the musician records themselves demonstrating a gesture, then trains a machine learning model to recognize and distinguish the target gesture from other body movements. Indeed, the NIME community has produced several software tools that facilitate this process \cite{bullock:mllib,fiebrink:wekinator, gillian:toolkit, caramiaux:nimeml, mcpherson:pipeline}.

These interactive machine learning tools are typically designed for offline training, wherein the musician records training data, trains the model, then tests it. With no knowledge of how the model will perform until after training, the musician must iterate, repeatedly recording, training, and testing until a satisfactory result is achieved. This opaque, iterative process can be a hindrance to creative flow, especially in live performance settings where improvising novel gesture-sound mappings might be desirable.

Ideally, one would be able to train a model in realtime, receiving live feedback on the model's performance as it learns each time a gesture is performed. Receving model performance reports on a per-gesture basis, the musician can build intuition for what the model has learned (e.g. which variations on the gesture are recognized and whicvh are not), adjust their gesture demonstrations accordingly, and arrive at a satisfactory result more efficiently.

This is the motivation behind \textit{Step-by-Step}, a software tool that allows users to train gesture recognition models with live audiovisual feedback. Here we present the system's hardware and software components, the method by which they are integrated, and preliminary results demonstrating the system's ability to learn to recognize and distinguish gestures in realtime.

\section{Background}
\subsection{Gesture Recognition}
Gesture recognition is the process of recognizing and distinguishing gestures in data. Gestures are body movement patterns that convey information. In the context of digital musical instruments, gestures are often used to control sound. For example, a musician might use a gesture to control the pitch of a note, the volume of a sound, or the timbre of an instrument.

@@

\subsection{Neural Networks}
Neural networks are one of many classes of machine learning models that can be used to analyze IMU timeseries and identify patterns such as gestures. In \textit{supervised learning}, neural networks are trained by providing them with examples of inputs and corresponding correct output labels. During training, the model adjusts its parameters to minimize the difference between its predictions and the groundtruth outputs. This process is called gradient descent and is performed iteratively until the model converges on a solution. Once trained, the model can be used to make predictions on novel inputs. Here, we perform gradient descent in parallel to the recording and automated labelling of training data, allowing the model to learn in real-time as new examples are provided.

@@

\begin{figure}[h]
    \includegraphics[width=0.5\textwidth]{../figures/model_diagram.jpg}
    \caption{Neural Network Architecture. The sensor branch (left) receives 20 frames (200 ms at 100 Hz) of 3-channel (x, y, z) IMU data. The sensor branch feeds into a 1D-convolutional layer with 8 filters, each also 20 frames in width. The memory branch (right) receives 15 frames of 1-hot encoded prediction data. The two branches are concatenated and fed into a fully connected output layer to predict the currently active gesture.}
    \label{fig:neural_network}
\end{figure}

@@

\begin{figure*}[h]
    \centering
    \includegraphics[width=1.0\textwidth]{../figures/realtime_training.jpg}
    \caption{Live Learning Feedback: The user performs a gesture and receives live feedback on the model's performance. Bottom Panel: 10 seconds of training data. Acceleration components (x, y, z) are shown in color; magnitude is shown in grey. Each magnitude peak is marked by a thin vertical line, indicating it has been labelled as a gesture in the training data. Top Panel: 10 seconds of inference data. At t = 0 seconds the neural network is completely untrained. Naive to what is and is not a gesture, it predicts gesture and non-gesture with equal probability (0.5). As the model trains on non-gesture date, it begins to predict non-gesture with higher probability. At t = 5 seconds, the model has converged on a solution that reliably distinguishes non-gesture. At t = 10 seconds, the model has converged on a solution that reliably distinguishes gestures. Given more training data and time to learn, the model's performance would continue to improve with gessture and non-gesture probabilities approaching 1 and 0, respectively. See supplementary video for live demo in the github repo.}
    \label{fig:realtime_training}
\end{figure*}

@@

The current hardware system uses a microcontroller to stream IMU data to a computer for processing. However, given the small size of the model (default settings: ~2kb with ~500 parameters) and recent advances in embedded machine learning, it should be possible to run the entire system in realtime on a wireless, wearable microcontroller, including labelling, training, and inference. This would allow the system to be used in a wider range of applications, including those where a computer is not available, desirable, or practical.

@@

\section{Conclusion}
Step-by-Step serves as a proof-of-concept for realtime learning of user-defined gestures in IMU data. It is worth noting that the strategies employed here are not limited to IMU data and could be applied to data from other sensors. The authors hope that this work will inspire others to explore the possibilities of realtime machine learning in human-computer interaction.

% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{nime-references}

%%% Place this command where you want to balance the columns on the last page. 
%\balancecolumns

% That's all folks!
\end{document}
